---
title: "Crop trade analysis"
author: "Robin Lindström"
date: "20/10/2019"
output:
  html_document:
      number_sections: true
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(root.dir = "~/Google Drive/SRC/Thesis/")
Sys.setlocale('LC_ALL','C')


library(scales)
library(gridExtra)
library(sf) # For loading map data
#library(tmap)
library(plotly)
#theme_set(theme_light(base_size = 8))
setwd(getwd())
```


```{r read_functions, echo = FALSE, include = FALSE}
source("~/Google Drive/Skola/SRC/Thesis/Code/Scripts/flex_crops_functions.R")
#setwd("~/Google Drive/SRC/Thesis/")
```

```{r get_raw_data, echo = FALSE, include = FALSE}
# Read the trade data

trade_data_path = "~/Google Drive/Skola/SRC/Thesis/Data/Trade data FAO/Trade_Crops_Livestock_E_All_Data_NOFLAG.csv"

export_data_raw = read_trade_data(trade_data_path) %>% 
  filter(str_detect(measures, "Export")) %>% 
  filter(crop_category != "Non crops")

import_data_raw = read_trade_data(trade_data_path) %>% 
  filter(str_detect(measures, "Import")) %>% 
  filter(crop_category != "Non crops")


```


```{r variables_and_general_data, echo = FALSE, include = FALSE}
years = 2010:2016
all_years = 1961:2016
years = all_years

all_years_columns = paste("y",all_years, sep = "")
year_columns = paste("y",years, sep = "")

remove_years = setdiff(all_years_columns, year_columns)

# Which columns am I interested in
selected_columns = c("country", "iso3_code", "item", "source", "measures", "crop_category",
                     "year", "value", "unit", "yearly_national_export", "country_group")

FAO_codes = get_fao_codes()
country_metrics = get_country_metrics()

excluded_crop_categories = c("Nuts", "Mushrooms & berries", "Spices", "Fruits", "Vegetables")

```

```{r process_export_data, echo = FALSE, include = FALSE}
# Which columns am I interested in
selected_columns = c("country", "iso3_code", "item", "source_crop", "measures", "crop_category",
                     "year", "value", "unit", "yearly_national_export", "country_group")

# Gather years into one column, replace all NA values with 0
export_data_processed = export_data_raw %>% 
#  filter(item_group == "Crops Primary") %>% 
#  select(-remove_years) %>% 
  gather(key = "year", "value", year_columns) %>% 
  mutate(year = as.numeric(gsub("y", "", year))) %>% 
  filter(item != "Total Merchandise Trade") %>% 
  group_by(year, country, measures) %>% 
  mutate(value = replace_na(value, 0)) %>% 
  mutate(yearly_national_export = sum(value)) %>% 
  ungroup() %>% 
  select(selected_columns)

all_crop_categories = export_data_processed %>% distinct(crop_category) %>% pull()
used_crop_categories = setdiff(all_crop_categories, excluded_crop_categories)

# Filter out only the value of exports and add some countriy specific data
export_value = export_data_processed %>% 
  filter(unit == "1000 US$") %>% 
  spread(measures, value) %>% 
  clean_names() %>%
#  left_join(land_area, by = "iso3_code") %>% 
  left_join(country_metrics, by = "iso3_code") %>% 
  filter(!is.na(land_area)) %>%  
  filter(land_area > 5000) %>% 
  filter(crop_category %in% used_crop_categories)

# These countries are excluded since they have an area smaller than 5000 sqm.

```

```{r process_import_data, echo = FALSE, include = FALSE}
selected_columns = c("country", "iso3_code", "item", "source_crop", "measures", "crop_category",
                     "year", "value", "unit", "yearly_national_import", "country_group")
# Gather years into one column, replace all NA values with 0
import_data_processed = import_data_raw %>% 
#  filter(item_group == "Crops Primary") %>% 
#  select(-remove_years) %>% 
  gather(key = "year", "value", year_columns) %>% 
  mutate(year = as.numeric(gsub("y", "", year))) %>% 
  filter(item != "Total Merchandise Trade") %>% 
  group_by(year, country, measures) %>% 
  mutate(value = replace_na(value, 0)) %>% 
  mutate(yearly_national_import = sum(value)) %>% 
  ungroup() %>% 
  select(selected_columns)

all_crop_categories = import_data_processed %>% distinct(crop_category) %>% pull()
used_crop_categories = setdiff(all_crop_categories, excluded_crop_categories)

# Filter out only the value of exports and add some countriy specific data
import_value = import_data_processed %>% 
  filter(unit == "1000 US$") %>% 
  spread(measures, value) %>% 
  clean_names() %>%
#  left_join(land_area, by = "iso3_code") %>% 
  left_join(country_metrics, by = "iso3_code") %>% 
  filter(!is.na(land_area)) %>%  
  filter(land_area > 5000) %>% 
  filter(crop_category %in% used_crop_categories)

# These countries are excluded since they have an area smaller than 5000 sqm.

```

# Analysis
# What is the proportion of flex crops imported in each country and which flex crop is the most dominating
```{r}
national_crop_group_import_value = import_value %>% 
  filter(crop_category != "Non crops",
         year == 2016) %>% 
  select(country, iso3_code, item, source, crop_category, import_value) %>% 
  group_by(country, iso3_code, source, crop_category) %>%
  summarise(import_value = sum(import_value)) %>% 
  ungroup() %>% 
  group_by(country) %>% 
  mutate(national_crop_import_value = sum(import_value)) %>% 
  #ungroup() %>% 
  #group_by(country) %>% 
  mutate(top_crop_group_import = source[import_value == max(import_value)]) %>% 
  ungroup() %>% 
  group_by(country, crop_category) %>% 
  mutate(national_crop_category_import_value = sum(import_value),
         national_crop_import_value_proportion = national_crop_category_import_value/national_crop_import_value)
  
national_crop_group_import_value_proportion = national_crop_group_import_value %>% 
  filter(crop_category == "Flex crops") %>% 
  ungroup() %>% 
  group_by(country) %>% 
  #filter(national_crop_category_import_value != 0) %>% 
  top_n(1,import_value)
  #mutate(top_flex_crop_import = source[import_value == max(import_value)]) %>% 
  #distinct(iso3_code, national_crop_import_value_proportion, top_flex_crop_import)

```

# What is the proportion of flex crops exported in each country and which flex crop is the most dominating

```{r}

national_crop_group_export_value = export_value %>% 
  filter(crop_category != "Non crops",
         year == 2016) %>% 
  select(country, iso3_code, item, source, crop_category, export_value) %>% 
  group_by(country, iso3_code, source, crop_category) %>%
  summarise(export_value = sum(export_value)) %>% 
  ungroup() %>% 
  group_by(country) %>% 
  mutate(national_crop_export_value = sum(export_value)) %>% 
  #ungroup() %>% 
  #group_by(country) %>% 
  mutate(top_crop_group_export = source[export_value == max(export_value)]) %>% 
  ungroup() %>% 
  group_by(country, crop_category) %>%
  mutate(national_crop_category_export_value = sum(export_value),
         national_crop_export_value_proportion = national_crop_category_export_value/national_crop_export_value)

national_crop_group_export_value_proportion = national_crop_group_export_value %>% 
  filter(crop_category == "Flex crops") %>% 
  ungroup() %>% 
  group_by(country) %>% 
  #mutate(top_flex = max(export_value)) %>%
  #distinct(country, top_flex) %>% 
  #View()
  #filter(national_crop_category_export_value != 0) %>% 
  top_n(1,export_value)
  #mutate(top_flex_crop_export = ifelse(export_value == 0, "None", source[export_value == max(export_value)])) %>% 
  #distinct(iso3_code, national_crop_export_value_proportion, top_flex_crop_export)

```



```{r flex_crop_total_production, fig.height = 4, fig.width = 8, echo = FALSE}

# How big is the export share in each country and each year for each crop? ----

# Calculate total annual national exports and each items export share of that total
export_value_processed = export_value %>% 
  mutate(export_value = replace_na(export_value, 0)) %>% 
  group_by(country, year) %>%
  mutate(total_export = sum(export_value)) %>% 
  mutate(export_share = export_value / total_export) %>% 
  arrange(desc(export_share))

```



```{r, echo = FALSE}
# Which countries has the highest share of agricultural share of total exports in 2016? ----
df_export_value_national = export_value %>%
  distinct(country, iso3_code, agriculture_export_share) %>%
  arrange(desc(agriculture_export_share))

df_export_map_data = make_map_data(df_export_value_national)

df_export_map_data %>%
  ggplot(aes(fill = agriculture_export_share)) +
  geom_sf() +
  scale_fill_gradient2()
  labs(title = "Flex crops as proportion of total area harvested", subtitle = "Change in percentage points",
       fill = str_wrap("Percentage points",5))

```

```{r global_crop_export_analysis, fig.height = 3, fig.width = 8, echo = FALSE}
theme_set(theme_light(base_size = 10))

# Make a plot over the global export value of different crop groups
export_commodity_value = export_value %>% 
  group_by(year, crop_category) %>% 
  summarise(total_crop_category_value_per_year = sum(export_value)) %>% 
  ungroup() %>% 
  group_by(year) %>% 
  mutate(total_value_per_year = sum(total_crop_category_value_per_year)) %>% 
  mutate(crop_category_share = total_crop_category_value_per_year / total_value_per_year)

# Plot the relative and absolute temporal change in export value of the different crop groups ---
theme_set(theme_bw(base_size = 8))
p1 = export_commodity_value %>% 
  ggplot(aes(x = year, y = crop_category_share, fill = crop_category)) +
  geom_area() +
  labs(title = "Crop group export value", subtitle = "Relative", x = "", y = "proportion", color = "") +
  scale_fill_brewer(palette = "Set2")

p2 = export_commodity_value %>% 
  ggplot(aes(x = year, y = total_crop_category_value_per_year,
             color = crop_category)) +
  geom_line() +
  labs(title = "Crop group export value",subtitle =  "Absolute", x = "", y = "export value ($)", color = "") +
  scale_color_brewer(palette = "Set2")

#ggarrange(p1, p2, ncol = 2)
grid.arrange(p1,p2, ncol = 2)

g <- arrangeGrob(p1, p2, ncol = 2) #generates g
suppressMessages(ggsave(file='images/export/global_crop_export_per_crop_group.png', g)) #saves g

```

```{r global_crop_import_analysis, fig.height = 3, fig.width = 8, echo = FALSE}
theme_set(theme_light(base_size = 10))

# Make a plot over the global import value of different crop groups
import_commodity_value = import_value %>% 
  group_by(year, crop_category) %>% 
  summarise(total_crop_category_value_per_year = sum(import_value)) %>% 
  ungroup() %>% 
  group_by(year) %>% 
  mutate(total_value_per_year = sum(total_crop_category_value_per_year)) %>% 
  mutate(crop_category_share = total_crop_category_value_per_year / total_value_per_year)

# Plot the relative and absolute temporal change in import value of the different crop groups ---
p1 = import_commodity_value %>% 
  ggplot(aes(x = year, y = crop_category_share, fill = crop_category)) +
  geom_area() +
  labs(title = "Crop group import value", subtitle = "Relative", x = "", y = "proportion", color = "") +
  scale_fill_brewer(palette = "Set2")

p2 = import_commodity_value %>% 
  ggplot(aes(x = year, y = total_crop_category_value_per_year,
             color = crop_category)) +
  geom_line() +
  labs(title = "Crop group import value", subtitle = "Absolute", x = "", y = "import value ($)", fill = "") +
  scale_color_brewer(palette = "Set2")

#ggarrange(p1, p2, ncol = 2)

grid.arrange(p1,p2, ncol = 2)

g <- arrangeGrob(p1, p2, ncol = 2) #generates g
suppressMessages(ggsave(file='images/export/global_crop_import_per_crop_group.png', g)) #saves g

```

```{r}
p1 = import_commodity_value %>% 
  ggplot(aes(x = year, y = total_crop_category_value_per_year,
             color = crop_category)) +
  geom_line() +
  labs(title = "Crop group import value", subtitle = "Absolute", x = "", y = "import value ($)", fill = "") +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~crop_category, scales = "free_y")

p2 = import_value %>% 
  filter(crop_category == "Flex crops") %>% 
  group_by(year, source) %>% 
  summarise(annual_individual_value = sum(import_value)) %>% 
  ggplot(aes(x = year, y = annual_individual_value, fill = source)) +
  geom_area() +
  labs(title = "Flex crops import value breakdown", y = "Import value (US$)", fill = "") +
  scale_fill_brewer(palette = "Set2")

p3 = import_value %>% 
  filter(crop_category == "Potential flex crops") %>% 
  group_by(year, source) %>% 
  summarise(annual_individual_value = sum(import_value)) %>% 
  ggplot(aes(x = year, y = annual_individual_value, fill = source)) +
  geom_area() +
  labs(title = "Potential flex crops import value breakdown", y = "Import value (US$)", fill = "") +
  scale_fill_brewer(palette = "Set2")

lay <- rbind(c(1,1),
               c(2,3))
  
#
grid.arrange(p1, p2, p3, layout_matrix = lay)

#save
g <- arrangeGrob(p1, p2, p3, layout_matrix = lay) #generates g
suppressMessages(ggsave(file='images/export/crop_group_import_breakdown.png', g)) #saves g

```

```{r}

lump_trade = function(lump_data, threshold = 0.8){
  #lump_data = other_commodity
  last_year = last(lump_data$year)
  plot_order = lump_data %>% 
      mutate(source = as.character(source)) %>%
      filter(year == last_year) %>% 
      arrange(desc(annual_individual_value)) %>% 
      mutate(rank = row_number(),
             cumulative_sum = cumsum(annual_individual_value),
             total_harvest = sum(annual_individual_value),
             perc = cumulative_sum / total_harvest)
  
  # Show all crops that together account for at least 90% of the volume, group the rest.
  # threshold = 0.8
  n_source = plot_order %>% filter(perc < threshold) %>% count(n()) %>% pull(n)
  
  lumped_data <- lump_data %>% 
      mutate(source = as.character(source)) %>%
      mutate(plot_label = ifelse(source %in% plot_order$source[1:n_source], source, 'Other')) %>%
      mutate(plot_label = factor(plot_label, levels = c((plot_order$source[1:n_source]), 'Other'))) %>%
      group_by(plot_label, year) %>%
      summarise(annual_individual_value = sum(annual_individual_value)) %>% 
      group_by(year) %>% 
      mutate(percentage = annual_individual_value / sum(annual_individual_value))
  return(lumped_data)
  }

other_commodity_crops = import_value %>% 
  filter(crop_category == "Other commodity crops") %>% 
  group_by(year, source) %>% 
  summarise(annual_individual_value = sum(import_value))

lumped_other_commodity_crops = lump_trade(other_commodity_crops, threshold = 0.8)

lumped_other_commodity_crops %>% 
  ggplot(aes(x = year, y = annual_individual_value, fill = plot_label)) +
  geom_area() +
  labs(title = "Other commodity crops import value breakdown", y = "Import value (US$)", fill = "") +
  scale_fill_brewer(palette = "Set2")

suppressMessages(ggsave(file='images/export/other_commodity_crops_import_breakdown.png')) #saves g


```

```{r}
other_oil_seeds = import_value %>% 
  filter(crop_category == "Other oil seeds") %>% 
  group_by(year, source) %>% 
  summarise(annual_individual_value = sum(import_value))

lumped_other_oil_seeds = lump_trade(other_oil_seeds, threshold = 0.8)

lumped_other_oil_seeds %>% 
  ggplot(aes(x = year, y = annual_individual_value, fill = plot_label)) +
  geom_area() +
  labs(title = "Other oil seeds import value breakdown", y = "Import value (US$)", fill = "") +
  scale_fill_brewer(palette = "Set2")

suppressMessages(ggsave(file='images/export/other_oil_seeds_import_breakdown.png')) #saves g
```

```{r}
other_cereals = import_value %>% 
  filter(crop_category == "Other cereals") %>% 
  group_by(year, source) %>% 
  summarise(annual_individual_value = sum(import_value))

lumped_other_cereals_crops = lump_trade(other_cereals, threshold = 0.8)

lumped_other_cereals_crops %>% 
  ggplot(aes(x = year, y = annual_individual_value, fill = plot_label)) +
  geom_area() +
  labs(title = "Other cereals import value breakdown", y = "Import value (US$)", fill = "") +
  scale_fill_brewer(palette = "Set2")

suppressMessages(ggsave(file='images/export/other_cereals_import_breakdown.png')) #saves g
```


```{r}
other_crops = import_value %>% 
  filter(crop_category == "Other crops") %>% 
  group_by(year, source) %>% 
  summarise(annual_individual_value = sum(import_value))

lumped_other_crops = lump_trade(other_crops, threshold = 0.6)

lumped_other_crops %>% 
  ggplot(aes(x = year, y = annual_individual_value, fill = plot_label)) +
  geom_area() +
  labs(title = "Other crops import value breakdown", y = "Import value (US$)", fill = "") +
  scale_fill_brewer(palette = "Set2")

suppressMessages(ggsave(file='images/export/other_crops_import_breakdown.png')) #saves g
```

```{r flex_crop_dominant_exports, echo = FALSE, include = FALSE}
# Check how many countries that has a flex crop as the most dominating crop

countries = sort((unique(export_value$country)))
n_countries = length(countries)
years = 1961:2016
n_years = length(years)
n_countries_with_flex_crop_dominating_export = rep(0, n_years)
#nations_dominate_flex_list = data.frame(10, 2010, 2011)

#nations_dominate_flex_list <- as.data.frame(matrix(0, ncol = n_years, nrow = n_countries))
#colnames(nations_dominate_flex_list) = paste0("y", years)
#rownames(nations_dominate_flex_list) = countries

nations_dominate_flex_empty_list = as.data.frame(expand.grid(countries, years)) %>% 
  rename(country = Var1, year = Var2)

temp = data.frame()

for(i in 1:n_years){
  nations_dominate_flex = export_value %>% 
    filter(year == years[i]) %>% 
    group_by(country, year) %>%
    mutate(yearly_national_export = sum(export_value)) %>% 
    ungroup() %>%
    group_by(country, year, source) %>%
    summarise(source_crop_export_share = sum(export_value) / min(yearly_national_export),
              crop_category = min(crop_category)) %>% 
    group_by(year, country) %>%
    top_n(1, source_crop_export_share) %>% 
    ungroup() %>% 
    select(country, year, source_crop_export_share, crop_category)
  
  temp = rbind(temp, nations_dominate_flex %>% 
                 select(-crop_category))

  #merge(nations_dominate_flex_list, nations_dominate_flex, by = c("country", "year"))
  
  current_value = nations_dominate_flex %>%
    filter(crop_category == "Flex crops") %>% 
    select(-crop_category) %>% 
    summarise(n_countries_with_flex_crop_dominating_import = n()) %>% 
    pull()
  
  n_countries_with_flex_crop_dominating_export[i] = current_value
}

nations_dominate_flex_list = nations_dominate_flex_empty_list %>% 
    left_join(temp, by = c("country" = "country", "year" = "year"))

flex_crop_export_share = data.frame(year = 1961:2016, n_countries_with_flex_crop_dominating_export = n_countries_with_flex_crop_dominating_export)

```


```{r flex_crop_dominant_export_plot, include = FALSE}

flex_crop_export_share %>% 
  ggplot(aes(year, n_countries_with_flex_crop_dominating_export)) +
    geom_line() +
    labs(title = "Number of countries with flex crop as most imported crop", x = "", y = "Number fo countries")

```

```{r flex_crop_dominant_imports, echo = FALSE, include = FALSE}
# Check how many countries that has a flex crop as the most dominating crop

countries = sort((unique(import_value$country)))
n_countries = length(countries)
years = 1961:2016
n_years = length(years)
n_countries_with_flex_crop_dominating_import = rep(0, n_years)
#nations_dominate_flex_list = data.frame(10, 2010, 2011)

#nations_dominate_flex_list <- as.data.frame(matrix(0, ncol = n_years, nrow = n_countries))
#colnames(nations_dominate_flex_list) = paste0("y", years)
#rownames(nations_dominate_flex_list) = countries

nations_dominate_flex_empty_list = as.data.frame(expand.grid(countries, years)) %>% 
  rename(country = Var1, year = Var2)

temp = data.frame()

for(i in 1:n_years){
  nations_dominate_flex = import_value %>% 
    filter(year == years[i]) %>% 
    group_by(country, year) %>%
    mutate(yearly_national_import = sum(import_value)) %>% 
    ungroup() %>%
    group_by(country, year, source) %>%
    summarise(source_crop_import_share = sum(import_value) / min(yearly_national_import),
              crop_category = min(crop_category)) %>% 
    group_by(year, country) %>%
    top_n(1, source_crop_import_share) %>% 
    ungroup() %>% 
    select(country, year, source_crop_import_share, crop_category)
  
  temp = rbind(temp, nations_dominate_flex %>% 
                 select(-crop_category))

  #merge(nations_dominate_flex_list, nations_dominate_flex, by = c("country", "year"))
  
  current_value = nations_dominate_flex %>%
    filter(crop_category == "Flex crops") %>% 
    select(-crop_category) %>% 
    summarise(n_countries_with_flex_crop_dominating_import = n()) %>% 
    pull()
  
  n_countries_with_flex_crop_dominating_import[i] = current_value
}

nations_dominate_flex_list = nations_dominate_flex_empty_list %>% 
    left_join(temp, by = c("country" = "country", "year" = "year"))

flex_crop_import_share = data.frame(year = 1961:2016, n_countries_with_flex_crop_dominating_import = n_countries_with_flex_crop_dominating_import)
```


```{r flex_crop_dominant_imports_plot, echo = FALSE, include = FALSE}
flex_crop_import_share %>% 
  ggplot(aes(year, n_countries_with_flex_crop_dominating_import)) +
    geom_line() +
    labs(title = "Number of countries with flex crop as most imported crop", x = "", y = "Number fo countries")
```


```{r flex_crop_dominant_trade_plot, echo = FALSE}

flex_crop_export_share %>% 
  left_join(flex_crop_import_share, by = "year") %>%
  rename(export = n_countries_with_flex_crop_dominating_export,import = n_countries_with_flex_crop_dominating_import) %>%
  gather(trade_type, n_countries, c(export, import)) %>% 
  ggplot(aes(year, n_countries, color = trade_type)) +
  geom_line() +
  labs(title = "Number of countries with flex crop as highest share of trade", x = "", y = "Number of countries",
       color = "Trade type") +
  theme_bw(base_size = 10) +
  scale_color_brewer(palette = "Set1")
  
suppressMessages(ggsave(file='images/export/no_countries_with_flex_crop_most_traded.png')) #saves g


```


# Find the change in exports (this is probably a bad way, do time series analysis instead)
```{r, echo = FALSE, include = FALSE}
# How much of total exports are 1, 3 and 5 crops?
countries_df = export_value %>%
  distinct(country, iso3_code) %>% 
  arrange(country)

variable <- sym(paste0("export_share_top", i, "_crops"))

crop_value_2016 = export_value %>% 
  filter(year == 2016) %>%
  filter(crop_category != "Non crops") %>% 
  mutate(individual_crop_proportion_exports = export_value / yearly_national_export) %>% 
  group_by(year, iso3_code, country, source) %>% 
  summarise(commodity_export = sum(export_value / yearly_national_export),
            crop_category = min(crop_category))

crop_value_1961 = export_value %>% 
  filter(year == 1961) %>%
  filter(crop_category != "Non crops") %>% 
  mutate(individual_crop_proportion_exports = export_value / yearly_national_export) %>% 
  group_by(year, iso3_code, country, source) %>% 
  summarise(commodity_export = sum(export_value / yearly_national_export),
            crop_category = min(crop_category))

export_share_2016 = crop_value_2016 %>% 
  top_n(1, commodity_export) %>% 
  ungroup() %>%
  group_by(iso3_code) %>% 
  summarise(export_share_2016 = sum(commodity_export))

export_share_1961 = crop_value_1961 %>% 
  top_n(1, commodity_export) %>% 
  ungroup() %>%
  group_by(iso3_code) %>% 
  summarise(export_share_1961 = sum(commodity_export))

national_crop_export_proportion = countries_df %>%
  left_join(export_share_2016, by = "iso3_code") %>%
  left_join(export_share_1961, by = "iso3_code") %>% 
  mutate(change = export_share_2016 - export_share_1961) %>%
  left_join(country_metrics, by = "iso3_code") %>% 
  left_join(FAO_codes, by = "iso3_code")
  #gather("type", "value", -c(country, iso3_code))

# countries_df %>%
#   gather("number_of_exporters", "value", -c(country, iso3_code)) %>% 
#   ggplot(aes(x = number_of_exporters, y = value)) +
#   geom_violin() +
#   geom_boxplot(width=0.1)
# 
# countries_df %>%
#   gather("number_of_exporters", "value", -c(country, iso3_code)) %>% 
#   ggplot(aes(value, fill = number_of_exporters)) +
#   #geom_histogram(bins = 20) +
#   geom_density(alpha=0.3)
```

# How has the national top exported crop share changed from 1961 to 2016?
```{r top_crop_export_share, echo = FALSE}

#FAO_codes = get_fao_codes() %>%
#  select(-country_code)
#country_metrics = get_country_metrics()


national_crop_export_proportion %>% 
  mutate(change_sign = change / abs(change)) %>% 
  mutate(limit = ifelse(export_share_2016 > 0.1, "high", "low")) %>% 
  filter(!is.na(limit)) %>%
  ggplot(aes(x = export_share_1961, y = export_share_2016, label = iso3_code, color = income)) + 
  geom_point() +
  geom_abline() +
  geom_text(hjust=0, vjust=0) +
  facet_wrap(~limit, scales = "free_y", ncol = 1)

  # ggplot(aes(x = export_share_1961, y = export_share_2016, label = iso3_code, color = income)) +
  # geom_point() +
  #   
  # geom_text(data=subset(national_crop_export_proportion,
  #                       export_share_1961 > 0.25 | export_share_2016 > 0.25), hjust=0, vjust=0) +
  # facet_wrap(~limit)
```

# Flex crops change over time

```{r, echo = FALSE, include = FALSE}
flex_crop_export_share_2016 = export_value %>% 
  filter(year == 2016) %>%
  filter(crop_category == "Flex crops") %>% 
  mutate(individual_crop_proportion_exports = export_value / yearly_national_export) %>% 
  group_by(iso3_code) %>% 
  summarise(flex_crop_export_share_2016 = sum(export_value / yearly_national_export))

flex_crop_export_share_1961 = export_value %>% 
  filter(year == 1961) %>%
  filter(crop_category == "Flex crops") %>% 
  mutate(individual_crop_proportion_exports = export_value / yearly_national_export) %>% 
  group_by(iso3_code) %>% 
  summarise(flex_crop_export_share_1961 = sum(export_value / yearly_national_export))

national_crop_export_proportion = countries_df %>%
  left_join(flex_crop_export_share_2016, by = "iso3_code") %>%
  left_join(flex_crop_export_share_1961, by = "iso3_code") %>% 
  mutate(change = flex_crop_export_share_2016 - flex_crop_export_share_1961) %>%
  left_join(country_metrics, by = "iso3_code") %>% 
  left_join(FAO_codes, by = "iso3_code")
  #gather("type", "value", -c(country, iso3_code))
```


```{r fig.height = 6, fig.width = 12}
#theme_set()
national_crop_export_proportion %>% 
  mutate(change_sign = change / abs(change)) %>% 
  filter(!is.na(change_sign)) %>% 
  mutate(limit = ifelse(flex_crop_export_share_2016 < 0.1 & flex_crop_export_share_1961 < 0.1, "low", "high")) %>% 
  filter(!is.na(limit),
         limit == "high") %>%
  ggplot(aes(x = flex_crop_export_share_1961, y = flex_crop_export_share_2016, label = iso3_code, color = country_group)) + 
  geom_point() +
  geom_abline() +
  geom_text(size = 6, hjust=0, vjust=0) +
  #facet_wrap(~change_sign, scales = "free") +
  theme_bw(base_size = 15)
```




```{r flex_crop_export_share_2, fig.width = 10, fig.height = 10}
df_export_map_data = make_map_data(national_crop_export_proportion)

temp = df_export_map_data %>% 
  select(gdp, country_group)

# df_export_map_data %>%
#   ggplot(aes(fill = flex_crop_export_share_2016)) +
#   geom_sf() +
#   scale_fill_gradient2() +
#   labs(title = "Flex crops export share", fill = "") +
#   theme_bw(base_size = 10) +
#   facet_grid(~country_group, scales = "free_y")
#   #facet_wrap(~country_group, scales = "free") 

my_map %>%
  ggplot(aes(fill = gdp)) +
  geom_sf() +
  scale_fill_gradient2() +
  theme_bw(base_size = 10)
  #facet_grid(~country_group, scales = "free")
  #facet_wrap(~country_group, scales = "free") 

```

